\documentclass[11pt,letterpaper,sans]{moderncv}

\moderncvstyle{banking} % 'casual', 'classic', 'oldstyle' 'banking', 'fancy'
\moderncvcolor{red} % 'blue' (default), 'orange', 'green', 'red', 'purple', 'grey' and 'black'

\usepackage[scale=0.75]{geometry} % Reduce document margins
\addtolength{\voffset}{-3.5em}
\addtolength{\textheight}{7em}
\setlength{\hintscolumnwidth}{2cm} % width of dates column

\quote{``It's hard to do a really good job
on anything you don't think about in the shower.''
--- Paul Graham}

\newcommand{\entry}[6]%
{\cventry[0.85em]{#4}{#3}{#2}{#1}{#5}{#6}} % banking
\newcommand{\cvurl}[1]{\bgroup\color{blue}\normalfont\url{#1}\egroup}
\newcommand{\arxiv}[1]{\bgroup\color{blue}\normalfont\rmfamily%
\href{http://arxiv.org/abs/#1}{arXiv:#1}\egroup}

\begin{filecontents}{biblio.bib}
@article{ref:schur,
  author    = {Evan Chen},
  title     = {Schur-Concavity for Avoidance of Increasing Subsequences in Block-Ascending Permutations},
  journal   = {Electr. J. Comb.},
  volume    = {24},
  number    = {4},
  pages     = {P4.4},
  year      = {2017},
  url       = {http://www.combinatorics.org/ojs/index.php/eljc/article/view/v24i4p4},
}

@article{ref:hypercube,
  author    = {Evan Chen},
  title     = {Linear polychromatic colorings of hypercube faces},
  journal   = {Electr. J. Comb.},
  volume    = {25},
  number    = {1},
  pages     = {P1.2},
  year      = {2018},
  url       = {http://www.combinatorics.org/ojs/index.php/eljc/article/view/v25i1p2},
}

@article{ref:thirds,
  author    = {Evan Chen},
  title     = {A Family of Partially Ordered Sets with Small Balance Constant},
  journal   = {Electr. J. Comb.},
  volume    = {25},
  number    = {4},
  pages     = {P4.43},
  year      = {2018},
  url       = {http://www.combinatorics.org/ojs/index.php/eljc/article/view/v25i4p43},
}

@article{ref:RNUT,
	year={2015},
	journal={Research in Number Theory},
	eid={28},
	volume={1},
	number={1},
	doi={10.1007/s40993-015-0028-0},
	title={Linnik's theorem for Sato-Tate laws on elliptic curves with complex multiplication},
	url={http://dx.doi.org/10.1007/s40993-015-0028-0},
	publisher={Springer International Publishing},
	author={Chen, Evan and Park, Peter S. and Swaminathan, Ashvin A.},
	language={English}
}

@article{ref:benford,
    AUTHOR = {Chen, Evan and Park, Peter S. and Swaminathan, Ashvin A.},
     TITLE = {On logarithmically {B}enford {S}equences},
   JOURNAL = {Proc. Amer. Math. Soc.},
  FJOURNAL = {Proceedings of the American Mathematical Society},
    VOLUME = {144},
      YEAR = {2016},
    NUMBER = {11},
     PAGES = {4599--4608},
      ISSN = {0002-9939},
     CODEN = {PAMYAR},
   MRCLASS = {11F11 (11N05)},
  MRNUMBER = {3544512},
       DOI = {10.1090/proc/13112},
       URL = {http://dx.doi.org/10.1090/proc/13112},
}

@book {ref:EGMO,
    AUTHOR = {Chen, Evan},
     TITLE = {Euclidean geometry in mathematical olympiads},
    SERIES = {MAA Problem Books Series},
 PUBLISHER = {Mathematical Association of America, Washington, DC},
      YEAR = {2016},
     PAGES = {xv+311},
      % ISBN = {978-0-88385-839-4; 978-1-61444-411-4},
   MRCLASS = {00A07 (51M04 51N20)},
  MRNUMBER = {3467691},
}

@article {ref:usamo,
    AUTHOR = {Chen, Evan and Ensley, Doug},
     TITLE = {45{$^{th}$} {U}nited {S}tates of {A}merica {M}athematical
              {O}lympiad and 7{$^{th}$} {U}nited {S}tates of {A}merica
              {J}unior {M}athematical {O}lympiad},
   JOURNAL = {Math. Mag.},
  FJOURNAL = {Mathematics Magazine},
    VOLUME = {89},
      YEAR = {2016},
    NUMBER = {4},
     PAGES = {303--311},
      ISSN = {0025-570X},
     CODEN = {MAMGA8},
   MRCLASS = {Preliminary Data},
  MRNUMBER = {3552776},
       DOI = {10.4169/math.mag.89.4.303},
       URL = {http://dx.doi.org/10.4169/math.mag.89.4.303},
}


@article {ref:erdos1,
    AUTHOR = {Deshouillers, J. M. and Erd{\H{o}}s, Paul and S{\'a}rk{\"o}zi,
              A.},
     TITLE = {On additive bases},
   JOURNAL = {Acta Arith.},
  FJOURNAL = {Polska Akademia Nauk. Instytut Matematyczny. Acta Arithmetica},
    VOLUME = {30},
      YEAR = {1976},
    NUMBER = {2},
     PAGES = {121--132},
      ISSN = {0065-1036},
   MRCLASS = {10L05},
  MRNUMBER = {0417105},
MRREVIEWER = {S. Znam},
}

@article{ref:erdos2,
    AUTHOR = {Martinet, J. and Deshouillers, J. M. and Cohen, Henri},
     TITLE = {La fonction somme des diviseurs},
 BOOKTITLE = {S\'eminaire de {T}h\'eorie des {N}ombres, 1972--1973 ({U}niv.
              {B}ordeaux {I}, {T}alence), {E}xp. {N}o. 11},
     PAGES = {10},
 PUBLISHER = {Lab. Th\'eorie des Nombres, Centre Nat. Recherche Sci.,
              Talence},
      YEAR = {1973},
   MRCLASS = {10A20 (10H10)},
  MRNUMBER = {0392794},
MRREVIEWER = {E. Grosswald},
}

@article{ref:erdos3,
    AUTHOR = {Cohen, Henri and Rubinstein-Salzedo, Simon and Thorne, Frank},
     TITLE = {Identities for field extensions generalizing the
              {O}hno-{N}akagawa relations},
   JOURNAL = {Compos. Math.},
  FJOURNAL = {Compositio Mathematica},
    VOLUME = {151},
      YEAR = {2015},
    NUMBER = {11},
     PAGES = {2059--2075},
      ISSN = {0010-437X},
   MRCLASS = {11R21 (11R29)},
  MRNUMBER = {3427573},
}

@article {ref:erdos4,
    AUTHOR = {Rubinstein-Salzedo, Simon and Swaminathan, Ashvin},
     TITLE = {Analysis on surreal numbers},
   JOURNAL = {J. Log. Anal.},
  FJOURNAL = {Journal of Logic and Analysis},
    VOLUME = {6},
      YEAR = {2014},
     PAGES = {Paper 5, 39},
      ISSN = {1759-9008},
   MRCLASS = {26E35 (03E10 54J05)},
  MRNUMBER = {3267278},
MRREVIEWER = {Philip Ehrlich},
}


@article{ref:ecexplicit,
    AUTHOR = {Chen, Evan and Park, Peter S. and Swaminathan, Ashvin A.},
     TITLE = {Elliptic curve variants of the least quadratic nonresidue
              problem and {L}innik's theorem},
   JOURNAL = {Int. J. Number Theory},
  FJOURNAL = {International Journal of Number Theory},
    VOLUME = {14},
      YEAR = {2018},
    NUMBER = {1},
     PAGES = {255--288},
      ISSN = {1793-0421},
   MRCLASS = {11G05 (11M41 11N05 11R44 14H52)},
  MRNUMBER = {3726253},
       URL = {https://doi.org/10.1142/S1793042118500161},
}

@article{ref:algint,
    AUTHOR = {Chen, Evan},
     TITLE = {Avoiding algebraic integers of bounded house in orbits of rational functions over cyclotomic closures},
   JOURNAL = {Proc. Amer. Math. Soc.},
  FJOURNAL = {Proceedings of the American Mathematical Society},
    VOLUME = {146},
      YEAR = {2018},
     PAGES = {4189-4198},
}

@article{ref:dmtcswilf,
  TITLE = {{The 26 Wilf-equivalence classes of length five quasi-consecutive
  patterns}},
  AUTHOR = {Chen, Evan and Narayanan, Shyam},
  URL = {https://dmtcs.episciences.org/4912},
  DOI = {},
  JOURNAL = {{Discrete Mathematics \& Theoretical Computer Science}},
  VOLUME = {{vol. 20 no. 2}},
  YEAR = {2018},
  MONTH = Oct,
  KEYWORDS = {Mathematics - Combinatorics ; 05A05, 05A19 ; G.2.1},
}

@article{ref:orthomorphism,
  title = {Multiplicative and exponential variations of orthomorphisms of cyclic groups},
  author = {Chen, Evan},
  url = {https://www.intlpress.com/site/pub/pages/journals/items/joc/content/vols/0011/0002/a008/index.php},
  doi = {10.4310/JOC.2020.v11.n2.a8},
  journal = {Journal of Combinatorics},
  volume = {11},
  number = {2},
  pages = {377 - 389},
  year = {2020},
  month = Jan,
}
\end{filecontents}

\usepackage[style=chem-rsc,maxnames=7]{biblatex}
\addbibresource{biblio.bib}

\lfoot{\emph{Last updated \today}}

\newcommand{\subtext}[1]{\bgroup\footnotesize\rmfamily\color{black!70}#1\egroup}
\newcommand{\previous}[1]{\bgroup\color{black}\bfseries#1\egroup}

\newcommand{\job}[4]{\smallskip\cvitemwithcomment{#2}{#1}{#3}%
\unskip\subtext{#4}\medskip}

\newcommand{\pub}[2]{\cvlistitem{#1. \\ \subtext{#2}}\smallskip}
\newcommand{\paper}[3]{\pub{#1}{#2. \arxiv{#3}.}}

\name{Evan}{Chen}
\email{evan@evanchen.cc}
\homepage{web.evanchen.cc} % URL
\social[github]{vEnhance}
\social[linkedin]{evanchenmath}
\social[facebook]{evanchenmath}
\lfoot{\emph{Last updated \today}}

\begin{document}

\makecvtitle % Print the CV title
\vspace*{-1.5em}

\section{Education}
\entry{Fall 2018 -- present}
	{Massachusetts Institute of Technology}%
	{PhD candidate in mathematics.}
	{}{}{}
(Full list of major coursework at
\cvurl{https://web.evanchen.cc/upload/math-coursework.pdf}.)

\subsection{Undergraduate}
\cvitemwithcomment{Massachusetts Institute of Technology}%
	{B.S.\ Math (transfer)}{2015 -- 2018}
\cvitemwithcomment{Harvard College}%
	{Undergraduate (and took math 55)}{2014 -- 2015}
\cvitemwithcomment{UC Berkeley}
	{Concurrent enrollment during high school}
	{2012 -- 2013}
\cvitemwithcomment{San Jos\'e State University}
	{Concurrent enrollment during high school}
	{2012 -- 2013}

\subsection{Programming experience}
Python (and Django), \LaTeX, Linux (Arch Linux on i3).

\section{Selected teaching and mentoring}
\job{Math Olympiad Summer Program (MOP)}%
{Assistant academic director}{2018 -- present}%
{Also \previous{head grader (2016-2017)}.
Training program for the USA team at the International Math Olympiad (IMO).
Gave lectures, coordinate grading, manage assistant teams,
direct logistics, exam design.
Reference: Po-Shen Loh.}

\job{USA Math Olympiad (USAMO)}%
{Co-Editor-in-Chief}{Starting July 2019}%
{Also \previous{grader (2015-), problem author (2016-), committee member (2018-2019)}.
Co-EiC with \emph{Jenny Iglesias} starting July 2019.
Reference: Bela Bajnok.}

\job{Olympiad Training for Individual Study (OTIS)}
{Director}{2015 -- present}
{Director of private individualized classes for students
who are preparing for USAMO/IMO type exams.
Details available at \cvurl{https://web.evanchen.cc/otis.html}.}

\job{USA Olympiad Team Selection Tests}%
{Team selection coordinator}{2017 -- present}%
{Together with \emph{Ankan Bhattacharya},
coordinating both the design and grading of team selection tests
for International Math Olympiad,
European Girls' Math Olympiad, and Romanian Masters in Mathematics.
Reference: Po-Shen Loh.}


\section{Selected fellowships and awards}
\cvitem{Morgan prize honorable mention}{awarded January 2019}
\cvitem{NSF graduate fellowship}{awarded April 2018}
\cvitem{MIT Akamai presidential graduate fellow}{awarded March 2018}

\cvitem{International Math Olympiad (IMO)}{Gold medal, awarded 2014, rank 12th}
\cvitem{USA Math Olympiad (USAMO)}{National winner, awarded 2014, rank 2nd}
\cvitem{William Lowell Putnam Competition}{N1, 2015, rank 13th}

\section{Publications and preprints}
\paper{A family of partially ordered sets with small balance constant}
{\fullcite{ref:thirds}.}{1709.05753}

\paper{Avoiding algebraic integers of bounded house in orbits of
rational functions over cyclotomic closures}
{\fullcite{ref:algint}}{1608.04146}

\paper{Elliptic curve variants of least quadratic nonresidue and Linnik's theorem}
{\fullcite{ref:ecexplicit}}{1507.07122}

\paper{Linear polychromatic colorings of hypercube faces}
{\fullcite{ref:hypercube}}{1609.01247}

\paper{Linnik's theorem for Sato-Tate laws on CM elliptic curves}%
{\fullcite{ref:RNUT}}{1506.09170}

\paper{Multiplicative and exponential orthomorphisms}
{\fullcite{ref:orthomorphism}}{1710.02734}

\paper{On logarithmically Benford sequences}%
{\fullcite{ref:benford}}{1507.02629} % July 2016

\paper{Schur-concavity for avoidance of increasing subsequences 
in block-ascending permutations}
{\fullcite{ref:schur}}{1708.01350}

\paper{The 26 Wilf-equivalence classes of length five quasi-consecutive patterns}
{\fullcite{ref:dmtcswilf}}{1609.04626}

\pub{Euclidean Geometry in Mathematical Olympiads}%
	{\fullcite{ref:EGMO}. ISBN 978-0-88385-839-4.%
	300 pp, 248 figures.
	Synopsis at \cvurl{https://web.evanchen.cc/geombook.html}.}

\pub{45th United States of America Mathematical Olympiad}{\fullcite{ref:usamo}.}

\section{Selected work experience}
\subsection{Research internships}
\job{Research Experience for Undergraduates}
{Duluth REU}{Summer 2016, 2017}
{Worked on projects in number theory and combinatorics.
Directed by Joe Gallian.}

\job{Research Experience for Undergraduates}
{Emory REU}{Summer 2015}
{Worked on projects in analytic number theory
on the distribution of primes. Directed by Ken Ono.}

\job{CEE Research Science Institute}
{RSI participant}{Summer 2013}
{Worked on high-school project on Bott-Samelson bimodules.
Directed by Center for Excellence in Education.}

\subsection{Volunteer and outreach}
\job{International Math Olympiad (IMO)}
{USA Observer A}{2017 -- present}
{Part of the USA delegation to the International Math Olympiad.
Advised on problem voting, helped organize trip for students,
and helped defend the papers of the USA team.}

\job{Harvard-MIT Math Tournament (HMMT)}
{Officer}{2014-2020}
{Served as \previous{problem czar (2014-2016),
software director (2016-2017), historian (2017-2018),
chief of staff (2018-2019), chief advisor (2019-2020)}.
See \cvurl{https://github.com/vEnhance/helium} for a software example.}

\job{American Regional Math League (ARML)}
{Problem writing team}{2016 -- present}
{Problem author team for the American Regional Math League
(ARML), an annual high-school math meet.}

\subsection{More teaching}
\job{MIT PRIMES}{Mentor and exam designer}{2016 -- present}%
{Was \previous{mentor (2016-2017), entrance exam designer (2016 - present)}.
Mentored for high-school/college collaborative research project
(PRIMES / AoPS CrowdMath); design project, oversee and guide discussion.
Reference: Tanya Khovanova.}

\job{Art of Problem Solving}{Assistant and curriculum designer}{2014 -- present}%
{Designed curriculum and materials, and worked as teaching assistant,
primarily for Worldwide Online Olympiad Training class (WOOT).
Reference: Richard Rusczyk.}

\job{Summer Program for Applied Rationality\&Cognition}%
{Counselor}{2015 -- present}
{Was \previous{junior counselor (2015, 2017, 2018),
admissions staff (2017, 2018).} Reference: Yan Zhang.}

\subsection{Software}
\job{Expii Inc.}{Software developer}{2013 -- 2015}
{Developer for expii.com, an online education platform.
Main task was the development and maintenance of the website's markup language,
both specifications and implementation.
Reference: Po-Shen Loh}

\end{document}
