\documentclass[11pt]{scrartcl}
\usepackage[sexy]{evan}
\usepackage{booktabs}
\usepackage{wrapfig}

\begin{document}
\title{Curves of degree greater than 2}
\subtitle{Some short notes from a MOP 2025 lecture}
\date{June 2025}
\maketitle

\setcounter{section}{-1}

\section{Warning}
This is a slight expansion of a \emph{in-person} lecture
I gave at the 2025 United States training camp.
That means there are going to be a \emph{lot} of details left out,
because the handout given was supposed to be just be a skeleton
for an actual lecture where the true dialogue was given orally.
I've gone back and added in some of the details for posterity,
but there are still a lot of unstated prerequisites.
In particular, I assume familiarity with $\mathbb{CP}^2$.

Also: to be clear, this handout is considered IMO-expert territory.
I can't blame students for being curious about fancy words
(really, curiosity is a good thing!), so here you go.
I just want to be clear that I consider this mostly for-fun side excursion,
rather than officially recommended training material for everyone.

\section{Cayley-Bacharach}
We work in $\mathbb{CP}^2$ always in this lecture.

\subsection{How many points determine a curve of degree $d$?}
For example, we all know that two points uniquely determine a line.
Algebraically, the way you can see this is that a line has coordinates
\[ Ax + By + C = 0 \]
for three constants $(A,B,C) \in \RR^3$, and two lines are the same if they have the same scale factor.
So you have $3-1 = 2$ degrees of freedom.

\dots At least, that's what I would say in $\RR^2$.
But actually, as I said, we're working in $\mathbb{CP}^2$.
So really I ought to write
\[ Ax + By + Cz = 0 \]
instead, with $(A,B,C) \in \CC^3$.
Note that we've made two changes:
we'll always work over $\CC$ instead of $\RR$,
and we use a homogeneous coordinate system $(x:y:z)$ rather than the Cartesian $(x,y)$.

For the same reason, we see five points ought to uniquely determine a conic:
\[ Ax^2 + Bxy + Cy^2 + Dxz + Eyz + Fz^2 = 0 \]
has six variables, so once we account for scaling, we get $6-1=5$ parameters.

More generally a degree-$d$ curve is given by a homogeneous polynomial of degree $d$ in
$\mathbb C[x,y,z]$, up to scaling.
This means a generic choice of $\binom{d+2}{2}-1$ points
should determine unique degree-$d$ curve: $2$ points determine a line,
$5$ points determine a conic, $9$ points determine a cubic, etc.
And for linear algebra reasons!

\subsection{Redundancy}
\dots Except the previous statement was a white lie;
it's only true in ``sufficiently general position''.
For example, if one takes four distinct collinear points,
any conic passing through three of them passes through the fourth.

Let's fix some (non-standard) terminology for this:
\begin{definition}
  Fix a degree $d \ge 0$.
  A set of points in $\mathbb{CP}^2$ is \emph{redundant} for degree-$d$ curves
  if vanishing on some proper subset implies vanishing on all of them.
\end{definition}
For example, three points are redundant for lines if and only if they're collinear.
(But four noncollinear points are redundant for lines too, due to vacuous truth ---
passing through $3$ points ``implies'' any conclusion.)

\subsection{Bezout's theorem}
The other fact about curves we should know is:
\begin{lemma}
  [Bezout theorem]
  Two curves of degree $a$ and $b$ sharing no irreducible component
  intersect at up to $ab$ points (in fact exactly $ab$ ``with multiplicity'').
\end{lemma}

\subsection{Chasles theorem for cubics}
Most of the time, we only see the cases $\max(a,b) \le 2$ in olympiad geometry,
and so Bezout theorem isn't that surprising.
However, let's think about the case $a=b=3$, where something weird first happens.

Let's say we draw two random irreducible cubics out of a hat, $\mathcal{C}_1 \neq \mathcal{C}_2$.
We expect them to meet at $9$ points.
However, we \emph{also} just said that $9$ points usually determine a unique cubic.
And yet we have two \emph{different} cubics passing through them!
So there must be some redundancy happening here.

That's what the following theorem makes precise.
According to \cite{EGH}, this should be called Chasles theorem (due to Michel Chasles).
\begin{theorem}
  [Chasles]
  Suppose two plane cubics meet at exactly $9$ points. Then:
  \begin{enumerate}[(a)]
    \itemsep1pt
    \ii This set of $9$ points is redundant for cubics.
    \ii No subset of size $8$ is redundant.
    \ii Hence, any cubic passing through any $8$ points must pass through the ninth.
  \end{enumerate}
\end{theorem}

As I just explained, part (a) is obvious:
cubics are supposed to be determined by $9$ points,
but we're \emph{given} two distinct cubics pass through these.
The bulk of the work of the theorem is to prove (b),
and it turns out you can do this with just Bezout, no algebraic geometry machinery:
\begin{problem}
  [{\cite[Proposition 1]{EGH}}]
  \label{prob:elem}
  Let $d \ge 1$ and consider $n$ points.
  Restrict attention
  to the case $n \le 2d+2$.
  Show that these points are redundant for degree-$d$ curves if and only if
  either $d+2$ are collinear, or $n = 2d+2$ and all points lie on a conic.
\end{problem}
I want to emphasize $n \le 2d+2$ is a huge assumption.
Removing this hypothesis is the main difficulty
to general Cayley-Bacharach statements.

\subsection{Application of Chasles theorem}
Part (c) of Chasles theorem is the one actually quoted by contestants,
and it follows from (a) and (b).
Here's an example of how you can use it:
\begin{example}
  [Pascal's theorem]
  Let $ABCDEF$ be a hexagon inscribed in a conic $\Gamma$.
  Then the points $X \coloneq \ol{BC} \cap \ol{EF}$,
  $Y \coloneq \ol{CD} \cap \ol{FA}$, $Z \coloneq \ol{DE} \cap \ol{AB}$ are collinear.
\end{example}

\begin{center}
  \begin{asy}
  size(6cm);

  draw(unitcircle);
  pair A = dir(85);
  pair B = dir(230);
  pair C = dir(55);
  pair D = dir(280);
  pair E = dir(135);
  pair F = dir(310);
  pair X = extension(B, C, E, F);
  pair Y = extension(C, D, F, A);
  pair Z = extension(D, E, A, B);

  draw(A--B, red);
  draw(C--D, red);
  draw(E--F, red);

  draw(B--C, blue);
  draw(D--E, blue);
  draw(F--A, blue);

  dot("$A$", A, dir(A));
  dot("$B$", B, dir(B));
  dot("$C$", C, dir(C));
  dot("$D$", D, dir(D));
  dot("$E$", E, dir(E));
  dot("$F$", F, dir(F));
  dot("$X$", X, dir(X));
  dot("$Y$", Y, dir(Y));
  dot("$Z$", Z, dir(Z));

  /* --------------------------------+
  | TSQX: by Evan Chen and CJ Quines |
  | https://github.com/vEnhance/tsqx |
  +----------------------------------+
  !size(6cm);

  unitcircle
  A = dir 85
  B = dir 230
  C = dir 55
  D = dir 280
  E = dir 135
  F = dir 310
  X = extension B C E F
  Y = extension C D F A
  Z = extension D E A B

  A--B / red
  C--D / red
  E--F / red

  B--C / blue
  D--E / blue
  F--A / blue
  */

  \end{asy}
\end{center}

\begin{proof}
  We consider the following three cubics:
  \begin{itemize}
    \ii Let $\mathcal{C}_1$ be the union of lines $AB$, $CD$, $EF$.
      (In the figure, these lines are red.)
    \ii Let $\mathcal{C}_2$ be the union of lines $BC$, $DE$, $FA$.
      (In the figure, these lines are blue.)
    \ii Let $\mathcal{C}_3$ be the union of line $XY$ and $\Gamma$.
  \end{itemize}
  Note that $\mathcal{C}_1$ and $\mathcal{C}_2$ meet at these nine points:
  \[ \mathcal{C}_1 \cap \mathcal{C}_2
    = \left\{ A, \; B, \; C, \; D, \; E, \; F, \; X, \; Y, \; Z \right\}. \]
  Also, $\mathcal{C}_3$ passes through the first eight points.
  So by part (c) of Chasles theorem, $\mathcal{C}_3$ passes through the ninth point $Z$ as well.
\end{proof}

\begin{remark}
  [Elliptic curve addition]
  For experts: another application of Chasles
  is to show that addition on an elliptic curve is associative.
  In fact one can even use this on cubic curves for competition problems
  --- by defining a group law on the cubics to gain additional collinearities.
\end{remark}

\subsection{Cayley-Bacharach for all degrees}
Higher-degree versions of Cayley-Bacharach come in different forms.
Here is one formulation that I've adapted for contest use from \cite[Theorem CB5]{EGH};
however, it doesn't handle multiplicity.
(There are versions of Cayley-Bacharach that do handle multiplicity,
but they require language about schemes to state correctly.)
\begin{theorem}
  [Cayley-Bacharach]
  Suppose two curves of degree $a$ and $b$ meet at exactly $ab$ points,
  partitioned into two sets $S \sqcup T$.
  Let $d,e \ge 0$ with $d+e = a+b-3$.
  Then the following statements are equivalent:
  \begin{enumerate}[(i)]
    \itemsep1pt
    \ii Any degree-$d$ curve passing through all of $S$ passes through all of $T$.
    \ii The set $T$ is not redundant for curves of degree $e$.
  \end{enumerate}
\end{theorem}
\noindent Special cases:
\begin{itemize}
  \itemsep1pt
  \ii Chasles theorem is the special case where $a=b=3$, $d=3$, $e=0$, $|S|=8$, $|T|=1$.

  \ii Another common case called ``\alert{quartic Cayley-Bacharach}''\footnote{The
    projective geometry chapter of \cite{MODA}
    claims that the following condition on $S$ also works
    (instead of $T$ non-collinear):
    no five points of $S$ are collinear,
    no nine are co-conic, no twelve are co-cubic.
    Proving this seems to need a stronger non-elementary version
    of \Cref{prob:elem} without the condition that $n \le 2d+2$.
    In any case, contestants are often sloppy with hypotheses.
    I was careful to show Chasles theorem with 3 parts,
    to emphasize that higher-degree versions are more subtle.}
  by contestants is when $a=b=4$,
  $d=4$, $e=1$, $|S|=13$, and $|T|=3$ (note this requires $T$ non-collinear).
  Then any quartic passing through the first $13$ points passes through the last $3$.
\end{itemize}

\section{Making Cayley-Bacharach practical for competing}
This is fine and all, but most olympiad problems you see won't have cubic curves
in their problem statements.
Therefore, in order to actually be useful, we need to be able to find a source of cubics.

In the example with Pascal's theorem,
we already saw how to make cubics by taking the unions of lines and circles.
That's the most common case!
To make this further practical, one will often use the \emph{circle points}
described shortly: these are two special points in $\mathbb{CP}^2$ that any two circles pass through.
(Remember, in $\mathbb{CP}^2$, we expect two circles to meet at $2 \cdot 2 = 4$ points.)

There are also two sources of \emph{irreducible} cubics: isopivotal and isoptic cubics.
We describe those too below briefly,
although we defer the example usage to the problems section.

\subsection{Circle points (most important and used all the time)}
In $\mathbb{CP}^2$, we consider two special points $\alpha$ and $\bar\alpha$
that every circle passes through, known as the \alert{circle points}.
In Cartesian coordinates they are defined via:
\begin{align*}
  \alpha &\coloneq (1:i:0) \\
  \bar{\alpha} &\coloneq (1:-i:0).
\end{align*}

\begin{definition}
  A cubic curve is \alert{circular} if it passes through $\alpha$ and $\bar\alpha$.
\end{definition}
\begin{exercise}
  Show that $\alpha$ and $\bar\alpha$ are isogonal conjugates
  with respect to any triangle.
  % The line at infinity swaps with the circumcircle, so it's clear
\end{exercise}
\begin{example}
  [Radical axis theorem]
  Show that the three common chords of any three pairwise intersecting circles concur.
\end{example}
\begin{proof}
  Refer to the following figure that I definitely did not steal from a
  textbook I wrote as a teenager.
  We call the three circles $\omega_1$, $\omega_2$, $\omega_3$.
  We let $\omega_2$ and $\omega_3$ meet at points $X$ and $X'$;
  define $Y$, $Y'$, $Z$, $Z'$ in the same way.
  \begin{center}
  \begin{asy}
    size(9cm);
    path w1 = circle((3,5), 4);
    path w2 = circle((0,0), 6);
    path w3 = circle((6,0), 5);
    pair[] X = intersectionpoints(w2, w3);
    pair[] Y = intersectionpoints(w3, w1);
    pair[] Z = intersectionpoints(w1, w2);
    draw(w1, gray);
    draw(w2, gray);
    draw(w3, gray);
    draw(X[0]--X[1]);
    draw(Y[0]--Y[1]);
    draw(Z[0]--Z[1]);
    dot("$X$", X[0], dir(X[0]-X[1]));
    dot("$X'$", X[1], dir(X[1]-X[0]));
    dot("$Y$", Y[0], dir(Y[0]-Y[1]));
    dot("$Y'$", Y[1], dir(Y[1]-Y[0]));
    dot("$Z$", Z[0], dir(Z[0]-Z[1]));
    dot("$Z'$", Z[1], dir(Z[1]-Z[0]));
    dot("$\alpha$ (on all circles)", (10,9), red);
    dot("$\bar{\alpha}$ (on all circles)", (10,7), red);
    label("$\omega_1$", (3,5)+4*dir(90), dir(90), mediumblue);
    label("$\omega_2$", (0,0)+6*dir(225), dir(225), mediumblue);
    label("$\omega_3$", (6,0)+5*dir(315), dir(315), mediumblue);
  \end{asy}
  \end{center}
  Again we consider three cubics:
  \begin{itemize}
    \ii Let $\mathcal{C}_1$ be the union of line $XX'$ with $\omega_1$.
    \ii Let $\mathcal{C}_2$ be the union of line $YY'$ with $\omega_2$.
    \ii Let $\mathcal{C}_3$ be the union of line $ZZ'$ with $\omega_3$.
  \end{itemize}
  Note that $\mathcal{C}_1$ and $\mathcal{C}_2$ meet at these nine points:
  \[ \mathcal{C}_1 \cap \mathcal{C}_2 = \left\{ X, \; X', \; Y, \; Y', \; Z, \; Z', \;
      \alpha, \; \bar\alpha, \; \ol{XX'} \cap \ol{YY'} \right\}. \]
  Also, $\mathcal{C}_3$ passes through the first eight points.
  So by part (c) of Chasles theorem, $\mathcal{C}_3$ passes through $\ol{XX'} \cap \ol{YY'}$ too.
\end{proof}


\subsection{Isopivotal cubics (used sometimes)}
In what follows $ABC$ is a fixed triangle and we use barycentric coordinates.
\begin{definition}
  Let $ABC$ be a triangle and $P = (u:v:w)$ a point.
  The \alert{isopivotal cubic} with pivot $P$ is the cubic curve
  \[ 0 = \det \begin{bmatrix} u & v & w \\
    x & y & z \\ a^2 yz & b^2 zx & c^2 xy \end{bmatrix}.  \]
  It contains all $X$ such that $X$ and its isogonal conjugate
  $X^\ast$ are collinear with $P$.
\end{definition}
\begin{proposition}
  [Points on isopivotal cubics]
  Every isopivotal cubic is self-isogonal and passes through all of the following $11$ points:
  the vertices $A$, $B$, $C$ of the triangle;
  the incenter and all three excenters;
  the pivot point $P$;
  and the points $AP \cap BC$, $BP \cap CA$, $CP \cap AB$.
\end{proposition}
\begin{exercise}
  For which $P$ is the isopivotal cubic with pivot $P$ a circular cubic?
\end{exercise}

Many choices of $P$ have names.
Most prominently:
\begin{itemize}
\ii The \emph{Thomson cubic} (also called the 23-point cubic) has pivot $G$ (centroid),
  and passes through $I$, $G$, $O$, $H$, $K$ and all altitude midpoints.
\ii The \emph{Neuberg cubic} (also called the 21-point or 37-point curve)
  has pivot $X_{30}$, the Euler infinity point (hence is circular),
  the reflections of $A$, $B$, $C$ across the opposite sides,
  and the six vertices of equilateral triangles constructed on the sides of $ABC$.
\ii The \emph{McCay cubic} has pivot $O$ (circumcenter);
  the \emph{orthocubic} has pivot $H$ (orthocenter);
  the \emph{Darboux cubic} has pivot at the de Longchamps point.
\end{itemize}

\subsection{Isoptic cubics (used sometimes)}
Now we assume $ABCD$ is a quadrilateral.
It could be self-intersecting, but we assume no three of vertices are collinear,
and $ABCD$ is not a parallelogram.
\begin{theorem}
  [{\cite[Theorem 1.5]{Hu}}]
  The following are equivalent for a point $P$:
  \begin{enumerate}[(i)]
    \itemsep1pt
    \ii $\dang APB = \dang DPC$.
    \ii $\angle APC$ and $\angle BPD$ have the same bisectors.
    \ii The feet from $P$ to the sides are concyclic.
    \ii $P$ has an isogonal conjugate with respect to $ABCD$.
    \ii $P$ is the focus of some ellipse tangent to the four sides.
  \end{enumerate}
\end{theorem}
\begin{theorem}
  [{\cite[Definition 2.3; Theorem 7.1]{Hu}}]
  The set of such $P$ is a nondegenerate \alert{circular} cubic curve
  passing through the vertices of $ABCD$,
  called the \alert{isoptic cubic}
  of $ABCD$.
  The isoptic cubic passes through the Miquel point $M$ of $ABCD$;
  moreover, $M\alpha$ and $M\bar\alpha$ are tangent to the isoptic cubic.
\end{theorem}
Note the emphasis on the word ``circular'' in the above theorem:
that's really important!
(Also, note that \cite{Hu} uses the name ``isogonal cubic'' instead.)

\begin{thebibliography}{9}
\bibitem[EGH95]{EGH}
  David Eisenbud, Mark Green, Joe Harris.
  \textit{Cayley-Bacharach theorems and conjectures.}

\bibitem[Gr12]{MODA}
  Adam Groucher. \textit{Mathematical Olympiad Dark Arts}.

\bibitem[Hu19]{Hu}
  Daniel Hu. \textit{Constructions in the locus of isogonal conjugates of a quadrilateral}.
  arXiv:1912.08296.
\end{thebibliography}

\newpage

\section{Problems}
\subsection{Problems for class}
These problems collectively cover the main techniques for the lecture
without being too complicated.
If this material is mostly new to you, these are good ones to start with.

% vanilla Chasles application
\begin{problem}
  [Miquel point]
  Triangle $DEF$ is inscribed in triangle $ABC$.
  Show that the circumcircles of $AEF$, $BFD$, $CDE$ meet at a point.
\end{problem}

% isoptic cubic
\begin{problem}
  [Romania TST 2017]
  Let $ABCD$ be a convex quadrilateral and let $P$ and $Q$ be variable points
  inside this quadrilateral so that $\angle APB=\angle CPD=\angle AQB=\angle CQD$.
  Prove that the lines $PQ$ obtained in this way all pass through a fixed point,
  or they are all parallel.
\end{problem}

% degree-four Cayley Bacharach
\begin{problem}
  % https://artofproblemsolving.com/community/c6h515448p2897891
  % Groucher
  Let $ABCD$ be a convex cyclic quadrilateral whose diagonals meet at $E$.
  Prove that there exists a conic passing through the incenters of triangles
  $ABC$, $BCD$, $CDA$, $DAB$, $ABE$, $BCE$, $CDE$, $DAE$.
\end{problem}

% isopivotal cubic using the infinity point along line $AE$
\begin{problem}[\href{https://aops.com/community/p27349354}{USAMO 2023/6}]
Let $ABC$ be a triangle with incenter $I$
and excenters $I_a$, $I_b$, $I_c$ opposite $A$, $B$, and $C$, respectively.
Given an arbitrary point $D$ on the circumcircle of $\triangle ABC$
that does not lie on any of the lines $II_a$, $I_bI_c$, or $BC$,
suppose the circumcircles of $\triangle DII_a$ and $\triangle DI_bI_c$
intersect at two distinct points $D$ and $F$.
If $E$ is the intersection of lines $DF$ and $BC$,
prove that $\angle BAD = \angle EAC$.
\end{problem}

\begin{wrapfigure}[12]{r}{6cm}
\centering
\begin{asy}
/*
Converted from GeoGebra by User:Azjps using Evan's magic cleaner
https://github.com/vEnhance/dotfiles/blob/main/py-scripts/export-ggb-clean-asy.py
*/
pair B = (10.12122,2.49871);
pair A = (2.99463,1.73330);
pair C = (11.69783,-4.81262);
pair D = (2.18048,-4.59861);
pair E = (4.32162,1.87582);
pair F = (8.72099,2.34833);
pair G = (10.46541,0.90259);
pair H = (11.18169,-2.41907);
pair I = (9.88187,-4.77178);
pair J = (3.81855,-4.63544);
pair K = (2.58210,-1.47503);
pair L = (2.69159,-0.62352);
pair W = (4.15065,-0.33709);
pair X = (9.00350,0.61559);
pair Y = (9.46760,-2.23090);
pair Z = (4.05027,-1.63620);

import graph;
size(5cm);
pen zzttqq = rgb(0.6,0.2,0.);
pen qqwuqq = rgb(0.,0.39215,0.);
pen cqcqcq = rgb(0.75294,0.75294,0.75294);
draw(A--B--C--D--cycle, linewidth(0.6) + zzttqq);

draw(circle((7.,-2.), 5.47544), linewidth(0.6) + blue);
draw(A--B, linewidth(0.6) + zzttqq);
draw(B--C, linewidth(0.6) + zzttqq);
draw(C--D, linewidth(0.6) + zzttqq);
draw(D--A, linewidth(0.6) + zzttqq);
draw(circle((6.92,-1.6), 4.33969), linewidth(0.6) + blue);
draw(F--I, linewidth(0.6));
draw(E--J, linewidth(0.6));
draw(L--G, linewidth(0.6));
draw(H--K, linewidth(0.6));
draw(circle((6.83965,-1.19828), 2.82354), linewidth(0.6) + dashed + qqwuqq);

dotfactor *= 2;
dot(A); dot(B); dot(C); dot(D);
dot(E); dot(F); dot(G); dot(H);
dot(I); dot(J); dot(K); dot(L);
dot(W); dot(X); dot(Y); dot(Z);
\end{asy}
\end{wrapfigure}

\subsection{Rumors}
The following problems have claimed solutions in various levels of detail,
but all of them have some part I'm not totally happy with.
If you're an expert in cubics, please help me fill in the gaps.

\begin{problem}[\href{https://aops.com/community/p23586650}{TSTST 2021/1}]
Let $ABCD$ be a quadrilateral inscribed in a circle with center $O$.
Points $X$ and $Y$ lie on sides $AB$ and $CD$, respectively.
Suppose the circumcircles of $ADX$ and $BCY$
meet line $XY$ again at $P$ and $Q$, respectively.
Show that $OP=OQ$.
\end{problem}

\begin{problem}
  [\href{https://danielphamnguyen.com/2025/01/30/generalized-cayley-bacharach-theorems/}{Folklore}]
  % see also https://geometriamafia.ru/material/2025/CB.pdf
  In the figure (with the outer two circles as given),
  show the four inner points are cyclic.
\end{problem}

\begin{problem}[\href{https://aops.com/community/p6580553}{TSTST 2016/6}]
Let $ABC$ be a triangle with incenter $I$, and whose incircle is tangent
to $\overline{BC}$, $\overline{CA}$, $\overline{AB}$ at $D$, $E$, $F$, respectively.
Let $K$ be the foot of the altitude from $D$ to $\overline{EF}$.
Suppose that the circumcircle of $\triangle AIB$
meets the incircle at two distinct points $C_1$ and $C_2$,
while the circumcircle of $\triangle AIC$ meets the incircle
at two distinct points $B_1$ and $B_2$.
Prove that the radical axis of the circumcircles of
$\triangle BB_1B_2$ and $\triangle CC_1C_2$ passes through the midpoint $M$ of $\overline{DK}$.
\end{problem}


\end{document}
